---
title: "Project 8 Template"
output: pdf_document
---

```{r}
# Add to this package list for additional SL algorithms
pacman::p_load(
  tidyverse,
  ggthemes,
  ltmle,
  tmle,
  SuperLearner,
  tidymodels,
  caret,
  dagitty,
  ggdag,
  here)

heart_disease <- read_csv(here('Projects/Project\ 8/heart_disease_tmle.csv'))
```

# Introduction

Heart disease is the leading cause of death in the United States, and treating it properly is an important public health goal. However, it is a complex disease with several different risk factors and potential treatments. Physicians typically recommend changes in diet, increased exercise, and/or medication to treat symptoms, but it is difficult to determine how effective any one of these factors is in treating the disease. In this project, you will explore SuperLearner, Targeted Maximum Likelihood Estimation (TMLE), and Longitudinal Targeted Maximum Likelihood Estimation (LTMLE). Using a simulated dataset, you will explore whether taking blood pressure medication reduces mortality risk.

# Data

This dataset was simulated using R (so it does not come from a previous study or other data source). It contains several variables:

```{=tex}
\begin{itemize}
    \item \textbf{blood\_pressure\_medication}: Treatment indicator for whether the individual took blood pressure medication (0 for control, 1 for treatment)
    \item \textbf{mortality}: Outcome indicator for whether the individual passed away from complications of heart disease (0 for no, 1 for yes)
    \item \textbf{age}: Age at time 1
    \item \textbf{sex\_at\_birth}: Sex assigned at birth (0 female, 1 male)
    \item \textbf{simplified\_race}: Simplified racial category. (1: White/Caucasian, 2: Black/African American, 3: Latinx, 4: Asian American, \newline 5: Mixed Race/Other)
    \item \textbf{income\_thousands}: Household income in thousands of dollars
    \item \textbf{college\_educ}: Indicator for college education (0 for no, 1 for yes)
    \item \textbf{bmi}: Body mass index (BMI)
    \item \textbf{chol}: Cholesterol level
    \item \textbf{blood\_pressure}: Systolic blood pressure 
    \item \textbf{bmi\_2}: BMI measured at time 2
    \item \textbf{chol\_2}: Cholesterol measured at time 2
    \item \textbf{blood\_pressure\_2}: BP measured at time 2
    \item \textbf{blood\_pressure\_medication\_2}: Whether the person took treatment at time period 2 
\end{itemize}
```
For the "SuperLearner" and "TMLE" portions, you can ignore any variable that ends in "\_2", we will reintroduce these for LTMLE.

# SuperLearner

## Modeling

Fit a SuperLearner model to estimate the probability of someone dying from complications of heart disease, conditional on treatment and the relevant covariates. Do the following:

```{=tex}
\begin{enumerate}
    \item Choose a library of at least 5 machine learning algorithms to evaluate. \textbf{Note}: We did not cover how to hyperparameter tune constituent algorithms within SuperLearner in lab, but you are free to do so if you like (though not required to for this exercise). 
    \item Split your data into train and test sets.
    \item Train SuperLearner
    \item Report the risk and coefficient associated with each model, and the performance of the discrete winner and SuperLearner ensemble
    \item Create a confusion matrix and report your overall accuracy, recall, and precision
\end{enumerate}
```
```{r}
head(heart_disease)

heart_disease_t1 <- heart_disease %>%
  select(-ends_with("_2"))

head(heart_disease_t1)
```

```{r}
listWrappers()
```

```{r}
# Fit SuperLearner Model

## sl lib 
## ^ select library after creating train/test split

## Train/Test split
heart_t1_split <- initial_split(heart_disease_t1, prop = 3/4)

heart_t1_train <- training(heart_t1_split)

heart_t1_test <- testing(heart_t1_split)

t1_y_train <- heart_t1_train %>%
  pull(mortality)

t1_y_test <- heart_t1_test %>%
  pull(mortality)

t1_x_train <- heart_t1_train %>%
  select(-mortality)

t1_x_test <- heart_t1_test %>%
  select(-mortality)

## Train SuperLearner
set.seed(123)
# Lasso
sl_group <- SuperLearner(Y = t1_y_train,              # target
                         X = t1_x_train,              # features
                         family = binomial(),      # binomial : 1,0s
                         SL.library = c("SL.glmnet", 
                                        "SL.mean", 
                                        "SL.knn",
                                        "SL.nnls",
                                        "SL.ranger")) # list models as a vector


```

```{r}

## Risk and Coefficient of each model
sl_group


```

```{r}
## Discrete winner and superlearner ensemble performance
sl_group$cvRisk[which.min(sl_group$cvRisk)]

## Confusion Matrix
```

```{r}
preds <- 
  predict(sl_group,             # use the superlearner not individual models
          t1_x_test,         # prediction on test set
          onlySL = TRUE)  # use only models that were found to be useful (had weights)


# start with y_test
validation <- 
  t1_y_test %>%
  # add our predictions - first column of predictions
  bind_cols(preds$pred[,1]) %>% 
  # rename columns
  rename(obs = `...1`,      # actual observations 
         pred = `...2`) %>% # predicted prob
  # change pred column so that obs above .5 are 1, otherwise 0
  mutate(pred = ifelse(pred >= .5, # this is the code that transforms the probabilities into binary
                           1,
                           0))

# view
head(validation)
```

```{r}
# confusion matrix
# ----------
caret::confusionMatrix(as.factor(validation$pred),
                       as.factor(validation$obs))
```

## Discussion Questions

```{=tex}
\begin{enumerate}
    \item Why should we, in general, prefer the SuperLearner ensemble to the discrete winner in cross-validation? Or in other words, what is the advantage of "blending" algorithms together and giving them each weights, rather than just using the single best algorithm (with best being defined as minimizing risk)?
\end{enumerate}
```
The choice of the "best" model in cross-validation can be sensitive to the particular sample of data used. Small changes in the training data can lead to different models being selected. The SuperLearner's ensemble method, by contrast, incorporates model averaging which can provide a more stable and reliable prediction by averaging across these variations. The SuperLearner method also allow for optimization across a variety of loss functions and is not tied to a single metric of success as cross-validation might be. This flexibility means it can be tailored to prioritize different aspects of prediction accuracy based on the specific context or application.

# Targeted Maximum Likelihood Estimation

## Causal Diagram

TMLE requires estimating two models:

```{=tex}
\begin{enumerate}
    \item The outcome model, or the relationship between the outcome and the treatment/predictors, $P(Y|(A,W)$.
    \item The propensity score model, or the relationship between assignment to treatment and predictors $P(A|W)$
\end{enumerate}
```
Using ggdag and daggity, draw a directed acylcic graph (DAG) that describes the relationships between the outcome, treatment, and covariates/predictors. Note, if you think there are covariates that are not related to other variables in the dataset, note this by either including them as freestanding nodes or by omitting them and noting omissions in your discussion.

```{r}
# DAG for TMLE
# install packages

# install.packages("ggdag")
# install.packages("dagitty")
# library(ggdag)
# library(dagitty)


```

```{r}
# Define the simplified DAG
# Define the DAG
dag <- dagitty::dagitty("dag {
  Covariates -> Treatment
  Covariates -> Outcome
  Unobserved -> Covariates
  Unobserved -> Treatment
  Treatment -> Outcome
}")

# Plot the DAG
ggdag_plot <- ggdag::ggdag(dag, text = TRUE)
print(ggdag_plot)

```

The DAG here shows covariates as a single node that are influenced by unobserved variables and themselves influence both treatment and outcome. I did not include any isolate covariates although I'm wondering if t2 treatments could be considered as such for this DAG. They wouldn't influence the outcome we have subset these data for and by definition cannot influence our t1 treatment.

## TMLE Estimation

Use the `tmle` package to estimate a model for the effect of blood pressure medication on the probability of mortality. Do the following:

```{=tex}
\begin{enumerate}
    \item Use the same SuperLearner library you defined earlier
    \item Use the same outcome model and propensity score model that you specified in the DAG above. If in your DAG you concluded that it is not possible to make a causal inference from this dataset, specify a simpler model and note your assumptions for this step.
    \item Report the average treatment effect and any other relevant statistics
\end{enumerate}
```
```{r}

head(heart_disease_t1)

Y <- heart_disease_t1 %>%
  pull(mortality)

A <- heart_disease_t1$blood_pressure_medication

W <- heart_disease_t1 %>%
  select(-c(blood_pressure_medication, mortality))
```

```{r}
sl_libs <- c('SL.mean', 'SL.ranger', 'SL.glm')
```

```{r}
set.seed(123)

# implement above all in one step using tmle
# ----------
tmle_fit <-
  tmle::tmle(Y = Y,                  # outcome
             A = A,                  # treatment
             W = W,                  # baseline covariates
             Q.SL.library = sl_libs, # libraries for initial estimate 
             g.SL.library = sl_libs) # libraries for prob to be in treatment

# view results 
tmle_fit
```

## Discussion Questions

```{=tex}
\begin{enumerate}
    \item What is a "double robust" estimator? Why does it provide a guarantee of consistency if either the outcome model or propensity score model is correctly specified? Or in other words, why does mispecifying one of the models not break the analysis? \textbf{Hint}: When answering this question, think about how your introductory statistics courses emphasized using theory to determine the correct outcome model, and in this course how we explored the benefits of matching.
\end{enumerate}
```
To be doubly robust requires either fitting the right model to estimate the expected outcome correctly or fitting the model to estimate the probability of treatment correctly. Having either one of these will still yield consistent estimations, i.e. the bias of the estimator will tend towards 0 as the sample size increases. If neither condition can be met then causal inference cannot be defined. If the propensity score model, which predicts the probability of receiving treatment based on covariates, is accurate, it addresses treatment assignment bias. Conversely, if the outcome model, which predicts outcomes based on treatment and covariates, is correct, it compensates for any inaccuracies in the treatment model.

# LTMLE Estimation

Now imagine that everything you measured up until now was in "time period 1". Some people either choose not to or otherwise lack access to medication in that time period, but do start taking the medication in time period 2. Imagine we measure covariates like BMI, blood pressure, and cholesterol at that time for everyone in the study (indicated by a "\_2" after the covariate name).

## Causal Diagram

Update your causal diagram to incorporate this new information. \textbf{Note}: If your groups divides up sections and someone is working on LTMLE separately from TMLE then just draw a causal diagram even if it does not match the one you specified above.

\textbf{Hint}: Check out slide 27 from Maya's lecture, or slides 15-17 from Dave's second slide deck in week 8 on matching.

\textbf{Hint}: Keep in mind that any of the variables that end in "\_2" are likely affected by both the previous covariates and the first treatment when drawing your DAG.

```{r}
# DAG for TMLE
dag <- dagitty::dagitty("dag {
  W1 -> A1 -> W2 -> A2 -> Y
  U -> W1
  U -> W2
  U -> A1
  U -> A2
  U -> Y
  W1 -> Y
  W1 -> A2
  W2 -> Y
  A1 -> Y
}")

# Plot the DAG
ggdag_plot <- ggdag::ggdag(dag, text = TRUE)
print(ggdag_plot)
```

## LTMLE Estimation

Use the `ltmle` package for this section. First fit a "naive model" that \textbf{does not} control for the time-dependent confounding. Then run a LTMLE model that does control for any time dependent confounding. Follow the same steps as in the TMLE section. Do you see a difference between the two estimates?\

```{r}
library(parallel)
library(SuperLearner)
library(tmle)
library(foreach)

# Setup parallel cluster
num_cores <- parallel::detectCores() - 2
cl <- makeCluster(num_cores)
doParallel::registerDoParallel(cl)
```

```{r}
colnames(heart_disease)
```

```{r}
# Randomly sampling a smaller size to hopefully prevent R from crashing
set.seed(123)
heart_disease_small <- heart_disease %>%
  select(c("blood_pressure", 
          "blood_pressure_medication",
          "blood_pressure_2",
          "blood_pressure_medication_2", 
          "mortality"))

heart_disease_small <- sample_frac(heart_disease_small, size = 0.10)

heart_disease_small_t1 <- heart_disease_small %>%
  select(-ends_with("_2"))
```

```{r}
## Naive Model (no time-dependent confounding) estimate
# implement tmle
# ----------
result <- ltmle(heart_disease_small_t1, 
                Anodes="blood_pressure_medication",       
                Lnodes=NULL,      # no Lnodes
                Ynodes="mortality",       
                abar=1,           
                SL.library=sl_libs)
# view
result



```

```{r}
colnames(heart_disease_small)
```

```{r}
## LTMLE estimate

result_ltmle <- ltmle(heart_disease_small, 
      Anodes=c("blood_pressure_medication", "blood_pressure_medication_2"),  # two treatment variables
      Lnodes=c("blood_pressure", "blood_pressure_2"),            # L indicator
      Ynodes="mortality",            # outcome
      abar=c(1, 1),          # treatment indicator in Anodes vector
      SL.library = sl_libs)

result_ltmle
```

## Discussion Questions

```{=tex}
\begin{enumerate}
    \item What sorts of time-dependent confounding should we be especially worried about? For instance, would we be concerned about a running variable for age the same way we might be concerned about blood pressure measured at two different times?
\end{enumerate}
```
Time-dependent confounding occurs when covariates affected by prior treatments influence future treatments and outcomes. For example, if blood pressure is lowered by medication and then influences further treatment decisions, it acts as a time-dependent confounder. Age, while continuously changing, typically does not change due to interventions and thus doesn't confound in the same way; however, its interaction with other variables and treatments can affect outcomes and should be controlled for.
